---
title: "Breast Cancer Spacedeconv Analysis (7 Slides)"
output:
  md_document:
    variant: gfm
---

```{r setup, message=FALSE, warning=FALSE}
library(spacedeconv)
library(ggplot2)
library(cowplot)

dir.create("../results/plots", recursive = TRUE, showWarnings = FALSE)

slides <- c("andersson", "1142243F", "1160920F", "4465", "44971", "4290", "4535")
title_size <- 18
font_size <- 14
legend_size <- 14
```

## Methods

Normalization and deconvolution follow the spacedeconv paper scripts; execution is implemented in `scripts/run_spacedeconv.R`.

- `preprocess(min_umi = 87)`
- `spacedeconv::normalize(...)`
- `deconvolute(method = "estimate", assay_sp = "cpm")`
- `deconvolute(method = "quantiseq", assay_sp = "cpm", tumor = TRUE)`
- `deconvolute(method = "epic", assay_sp = "cpm", tumor = TRUE)`

Rectangle integration is executed by:

- `scripts/export_spatial_for_rectangle.R`
- `scripts/run_rectangle.py`
- `scripts/merge_rectangle_results.R`

## Per-slide Spatial Results

```{r per-slide-plots, message=FALSE, warning=FALSE, fig.width=16, fig.height=6}
for (slide in slides) {
  cat("\n### Slide ", slide, "\n\n", sep = "")

  est <- readRDS(file.path("../results/objects", paste0("deconv_estimate_", slide, ".rds")))
  qua <- readRDS(file.path("../results/objects", paste0("deconv_quantiseq_", slide, ".rds")))
  epi <- readRDS(file.path("../results/objects", paste0("deconv_epic_", slide, ".rds")))

  p_tumor <- plot_spatial(
    est,
    result = "estimate_tumor.purity",
    smooth = TRUE,
    density = FALSE,
    title = paste0(slide, " - Tumor Purity (ESTIMATE)"),
    title_size = title_size,
    font_size = font_size,
    legend_size = legend_size
  )

  p_cd8 <- plot_spatial(
    qua,
    result = "quantiseq_T.cell.CD8.",
    smooth = TRUE,
    density = FALSE,
    shift_positive = FALSE,
    title = paste0(slide, " - CD8 T cells (quanTIseq)"),
    title_size = title_size,
    font_size = font_size,
    legend_size = legend_size
  )

  p_caf <- plot_spatial(
    epi,
    result = "epic_Cancer.associated.fibroblast",
    smooth = TRUE,
    density = FALSE,
    title = paste0(slide, " - CAFs (EPIC)"),
    title_size = title_size,
    font_size = font_size,
    legend_size = legend_size
  )

  rect_csv <- file.path("../results/objects", paste0("deconv_rectangle_", slide, ".csv"))
  if (file.exists(rect_csv)) {
    rect <- read.csv(rect_csv, stringsAsFactors = FALSE, check.names = FALSE)
    rect_cols <- setdiff(colnames(rect), "spot_id")

    if (length(rect_cols) >= 2) {
      colData(est)$rectangle_target_1 <- rect[[rect_cols[[1]]]][match(colnames(est), rect$spot_id)]
      colData(est)$rectangle_target_2 <- rect[[rect_cols[[2]]]][match(colnames(est), rect$spot_id)]

      p_rect_1 <- plot_spatial(
        est,
        result = "rectangle_target_1",
        smooth = TRUE,
        density = FALSE,
        shift_positive = FALSE,
        title = paste0(slide, " - Rectangle: ", rect_cols[[1]]),
        title_size = title_size,
        font_size = font_size,
        legend_size = legend_size
      )

      p_rect_2 <- plot_spatial(
        est,
        result = "rectangle_target_2",
        smooth = TRUE,
        density = FALSE,
        shift_positive = FALSE,
        title = paste0(slide, " - Rectangle: ", rect_cols[[2]]),
        title_size = title_size,
        font_size = font_size,
        legend_size = legend_size
      )

      panel <- cowplot::plot_grid(p_tumor, p_cd8, p_caf, p_rect_1, p_rect_2, ncol = 5, align = "hv")
    } else {
      panel <- cowplot::plot_grid(p_tumor, p_cd8, p_caf, ncol = 3, align = "hv")
    }
  } else {
    panel <- cowplot::plot_grid(p_tumor, p_cd8, p_caf, ncol = 3, align = "hv")
  }

  print(panel)

  ggsave(
    filename = file.path("../results/plots", paste0("slide_", slide, "_targets.png")),
    plot = panel,
    width = 20,
    height = 5,
    units = "in",
    dpi = 300,
    bg = "white"
  )
}
```

## Rectangle vs Existing Methods

```{r rectangle-comparison, message=FALSE, warning=FALSE}
tab <- read.csv("../results/tables/targets_all_slides.csv", stringsAsFactors = FALSE, check.names = FALSE)
rect_cols <- grep("^rectangle_", colnames(tab), value = TRUE)

if (length(rect_cols) == 0) {
  cat("Rectangle columns not found in combined table. Run merge script first.")
} else {
  pick_col <- function(patterns) {
    for (p in patterns) {
      hit <- grep(p, rect_cols, value = TRUE, ignore.case = TRUE)
      if (length(hit) > 0) {
        return(hit[[1]])
      }
    }
    NA_character_
  }

  rect_cd8 <- pick_col(c("cd8", "t.cell", "cytotoxic"))
  rect_caf <- pick_col(c("caf", "fibro", "fibroblast", "stromal"))

  out <- data.frame(
    metric = c("CD8 method correlation", "CAF method correlation"),
    rectangle_column = c(rect_cd8, rect_caf),
    correlation = c(NA_real_, NA_real_),
    stringsAsFactors = FALSE
  )

  if (!is.na(rect_cd8) && "cd8_quantiseq" %in% colnames(tab)) {
    out$correlation[out$metric == "CD8 method correlation"] <- suppressWarnings(
      cor(tab[[rect_cd8]], tab$cd8_quantiseq, use = "pairwise.complete.obs")
    )
  }

  if (!is.na(rect_caf) && "caf_epic" %in% colnames(tab)) {
    out$correlation[out$metric == "CAF method correlation"] <- suppressWarnings(
      cor(tab[[rect_caf]], tab$caf_epic, use = "pairwise.complete.obs")
    )
  }

  knitr::kable(out)
}
```

## Combined Table Preview

```{r combined-table, message=FALSE, warning=FALSE}
tab <- read.csv("../results/tables/targets_all_slides.csv")
knitr::kable(head(tab, 20))
```

## Session Info

```{r session-info}
sessionInfo()
```
