---
title: "Breast Cancer Architecture Bulk RNA-seq Analysis (TCGA-BRCA)"
output:
  md_document:
    variant: gfm
---

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(knitr)

dir.create("../results/bulk_rnaseq/plots", recursive = TRUE, showWarnings = FALSE)
raw_tab <- read.csv("../results/bulk_rnaseq/tables/tcga_brca_all_methods_raw.csv", check.names = FALSE)
tab <- read.csv("../results/bulk_rnaseq/tables/tcga_brca_all_methods_aggregated.csv", check.names = FALSE)
```

## Methods

- Input dataset: `data/bulk_rnaseq/tcga_brca/brca_tpm_all_symbols.csv`
- Samples processed: all samples in the TPM matrix
- 1st gen methods were run through `immunedeconv`:
  - `estimate`
  - `quantiseq`
  - `epic`
- 2nd gen method:
  - `Rectangle`
- Rectangle used the existing Wu reference and current `config/rectangle.yaml`

## Dataset Summary

```{r dataset-summary, message=FALSE, warning=FALSE}
summary_df <- data.frame(
  metric = c("n samples", "n columns raw merged", "n columns aggregated"),
  value = c(nrow(tab), ncol(raw_tab), ncol(tab))
)
kable(summary_df)
```

```{r metadata-summary, message=FALSE, warning=FALSE}
meta_cols <- intersect(c("sample_type", "PAM50"), colnames(tab))
if (length(meta_cols) == 0) {
  cat("No metadata columns available in aggregated table.")
} else {
  for (col in meta_cols) {
    cat("\n###", col, "\n\n")
    kable(head(sort(table(tab[[col]]), decreasing = TRUE), 20))
  }
}
```

## Raw Output Preview

```{r raw-preview, message=FALSE, warning=FALSE}
kable(head(raw_tab, 10))
```

## Aggregated Comparison

```{r comparison-summary, message=FALSE, warning=FALSE}
mk_scatter <- function(df, x, y, title, xlab, ylab, out_png) {
  d <- data.frame(x = as.numeric(df[[x]]), y = as.numeric(df[[y]]))
  d <- d[complete.cases(d), , drop = FALSE]
  r <- if (nrow(d) > 1) cor(d$x, d$y, method = "pearson") else NA_real_
  p <- ggplot(d, aes(x = x, y = y)) +
    geom_point(size = 0.6, alpha = 0.35) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(
      title = title,
      subtitle = paste0("Pearson r = ", ifelse(is.na(r), "NA", sprintf("%.3f", r))),
      x = xlab,
      y = ylab
    ) +
    theme_bw(base_size = 11)
  ggsave(out_png, p, width = 6, height = 5, dpi = 300, bg = "white")
  print(p)
  invisible(r)
}

comparisons <- data.frame(
  metric = c("Tumor", "CD8", "CAF", "B cells"),
  x = c("estimate_tumor", "quantiseq_cd8", "epic_caf", "epic_bcells"),
  y = c("rectangle_tumor", "rectangle_cd8", "rectangle_caf", "rectangle_bcells"),
  stringsAsFactors = FALSE
)

cors <- numeric(nrow(comparisons))
for (i in seq_len(nrow(comparisons))) {
  row <- comparisons[i, ]
  png_path <- file.path("../results/bulk_rnaseq/plots", paste0("tcga_brca_", make.names(row$metric), "_scatter.png"))
  cors[[i]] <- mk_scatter(
    tab, row$x, row$y,
    paste0("TCGA-BRCA ", row$metric, ": 1st-gen vs Rectangle"),
    paste0("1st-gen ", row$metric),
    paste0("Rectangle ", row$metric),
    png_path
  )
}

comparisons$pearson_r <- cors
kable(comparisons[, c("metric", "x", "y", "pearson_r")])
```

## Session Info

```{r session-info}
sessionInfo()
```
