---
title: "Breast Cancer Architecture Bulk RNA-seq Analysis (CPTAC-BRCA)"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(knitr)
library(jsonlite)

dir.create("../results/bulk_rnaseq/cptac_brca/plots", recursive = TRUE, showWarnings = FALSE)

raw_tab <- read.csv("../results/bulk_rnaseq/cptac_brca/tables/cptac_brca_all_methods_raw.csv", check.names = FALSE)
tab <- read.csv("../results/bulk_rnaseq/cptac_brca/tables/cptac_brca_all_methods_aggregated.csv", check.names = FALSE)
export_tab <- read.csv("../tum_export/deconvolution_results/CPTAC_BRCA_bulk_deconvolution_results_aggregated_rectangle_cell_types.csv", check.names = FALSE)
prep_diag <- fromJSON("../results/bulk_rnaseq/cptac_brca/intermediate/cptac_brca_input_summary.json")
rect_diag <- fromJSON("../results/bulk_rnaseq/cptac_brca/objects/deconv_rectangle_cptac_brca_diagnostics.json")
```

## Methods

- Input dataset: `data/bulk_rnaseq/cptac_brca/tpm_matrix_gene_id_name.csv`
- `gene_id` / Ensembl identifiers were removed from the analysis matrix.
- `gene_name` was used as the gene symbol identifier.
- Duplicated gene symbols were resolved by keeping the row with the highest mean TPM across all samples.
- 1st generation methods were run through `immunedeconv`:
  - `estimate`
  - `quantiseq`
  - `epic`
- 2nd generation method:
  - `Rectangle`
- Rectangle used the existing Wu reference and current `config/rectangle.yaml`.
- No CPTAC clinical/PAM50/survival metadata were joined in this workflow.

## Dataset Summary

```{r dataset-summary}
summary_df <- data.frame(
  metric = c(
    "n samples",
    "n input rows",
    "n unique gene symbols",
    "n duplicate rows removed",
    "n raw merged columns",
    "n aggregated columns",
    "n export columns",
    "Rectangle gene overlap",
    "Rectangle CPUs"
  ),
  value = c(
    prep_diag$n_samples,
    prep_diag$n_rows_input,
    prep_diag$n_rows_after_duplicate_resolution,
    prep_diag$n_duplicate_rows_removed,
    ncol(raw_tab),
    ncol(tab),
    ncol(export_tab),
    rect_diag$n_genes_overlap,
    rect_diag$n_cpus
  )
)
kable(summary_df)
```

## Output Preview

```{r output-preview}
kable(head(export_tab, 10))
```

## Method Comparison

```{r comparison-summary, fig.width=6, fig.height=5}
mk_scatter <- function(df, x, y, title, xlab, ylab, out_png) {
  d <- data.frame(x = as.numeric(df[[x]]), y = as.numeric(df[[y]]))
  d <- d[complete.cases(d), , drop = FALSE]
  r <- if (nrow(d) > 1) cor(d$x, d$y, method = "pearson") else NA_real_
  p <- ggplot(d, aes(x = x, y = y)) +
    geom_point(size = 1.4, alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(
      title = title,
      subtitle = paste0("Pearson r = ", ifelse(is.na(r), "NA", sprintf("%.3f", r))),
      x = xlab,
      y = ylab
    ) +
    theme_bw(base_size = 11)
  ggsave(out_png, p, width = 6, height = 5, dpi = 300, bg = "white")
  print(p)
  invisible(r)
}

comparisons <- data.frame(
  metric = c("Tumor", "CD8", "CAF", "B cells"),
  method = c("ESTIMATE", "quanTIseq", "EPIC", "EPIC"),
  x = c("estimate_tumor", "quantiseq_cd8", "epic_caf", "epic_bcells"),
  y = c("rectangle_tumor", "rectangle_T.cells.CD8.", "rectangle_caf", "rectangle_bcells"),
  stringsAsFactors = FALSE
)

cors <- numeric(nrow(comparisons))
for (i in seq_len(nrow(comparisons))) {
  row <- comparisons[i, ]
  png_path <- file.path("../results/bulk_rnaseq/cptac_brca/plots", paste0("cptac_brca_", make.names(row$metric), "_scatter.png"))
  cors[[i]] <- mk_scatter(
    tab, row$x, row$y,
    paste0("CPTAC-BRCA ", row$metric, ": ", row$method, " vs Rectangle"),
    paste0("1st-gen ", row$metric),
    paste0("Rectangle ", row$metric),
    png_path
  )
}
```

```{r correlation-table}
comparisons$pearson_r <- cors
kable(comparisons[, c("metric", "method", "x", "y", "pearson_r")])
```

## Rectangle Export Checks

```{r rectangle-export-checks}
agg_cols <- grep("^rectangle_aggregated_", colnames(export_tab), value = TRUE)
sum_with_removed <- rowSums(export_tab[, agg_cols, drop = FALSE], na.rm = TRUE) + tab[["rectangle_NKT.cells"]]
check_df <- data.frame(
  metric = c(
    "n rectangle aggregated export columns",
    "n fine rectangle export columns",
    "n old coarse rectangle export columns",
    "min aggregated + removed NKT",
    "max aggregated + removed NKT"
  ),
  value = c(
    length(agg_cols),
    sum(grepl("^rectangle_fine_", colnames(export_tab))),
    sum(colnames(export_tab) %in% c("rectangle_tumor", "rectangle_cd8", "rectangle_caf", "rectangle_bcells")),
    min(sum_with_removed, na.rm = TRUE),
    max(sum_with_removed, na.rm = TRUE)
  )
)
kable(check_df)
```

## Notes

EPIC reported incomplete optimization convergence for five CPTAC samples:
`05BR042`, `11BR036`, `01BR027`, `14BR005`, and `18BR004`.
Their `epic_caf` and `epic_bcells` values were within the observed CPTAC range, but these samples should be flagged for EPIC-specific interpretation.

## Session Info

```{r session-info}
sessionInfo()
```
